---
title: "chapter6"
author: "Ghaida Azzahra"
date: "2022-12-12"
output: html_document
---

# Assignment 6: Analysis of longitudinal data

```{r}
date()
```

## RATS data

### Exploring the data

```{r}
# read the original data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

# read the wrangled long data
RATSL <- read.csv(file = "ratsl.csv", sep=",", header=TRUE)

# Look at the (column) names 
names(RATSL)

# Look at the structure 
str(RATSL)

# Print out summaries of the variables
summary(RATSL)
```

This data is recorded for a nutrition study. The study involves three groups of rats, each were put on a different diet. Then, their weights (in grams) were recorded repeatedly to see if the diets affect the growth of their weights.

There are 176 observations in total and 5 variables, namely:

| Variable | Description                                        |
|----------|----------------------------------------------------|
| ID       | ID of the rat (integer, 1-16)                      |
| Group    | Group of the rat (integer, 1-3)                    |
| WD       | Measurement day (character, eleven categories)     |
| Weight   | Recorded weight of the rat in grams (integer)      |
| Time     | Measurement day (integer, eleven different values) |

### Factor the categorical variables

```{r, warning =FALSE, message=FALSE}
# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)
library(tidyverse)

# turn into factors
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# the data is already converted to the long form in the wrangling exercise

# checking 
glimpse(RATSL)
```

Variables ID and Group had been successfully turned into factor variables.

### Plotting

```{r, warning=FALSE, message=FALSE}
#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight))) 

```

From this plot, we can see that the rats are grouped according to their weights, although there is one rat that is significantly heavier than other rats in Group 2. The rats in the first group have the lowest weights compared to the other groups. Meanwhile, most of the rats in the second group are lighter than the one in the third group, although the heaviest rat of all groups is allocated the second group. The weight of the rats in all groups seem to increase over time, but it is still unclear yet if the growth is different in each group.

### Standardizing the data

```{r}
# Standardise the variable weight
RATSL <- RATSL %>%
  group_by(Group) %>%
  mutate(stdWeight = c(scale(Weight))) %>%
  ungroup()

# Plot again with the standardised data
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdWeight), max(RATSL$stdWeight))) 
```

After being standardized, the weight difference between the groups are not as pronounced anymore. Now, we can see the changes that happen in each group more clearly. Although it is still unclear if there are any differences between each group, lighter rats in all groups seem to have the most extreme weight gain compared to heavier rats in its group.

### Summary graph

```{r, warning=FALSE, message=FALSE}
# Number of Time
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of Weight by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
```

The plot above shows the average increase of weight over time in each group of rats. The second group seem to have the most pronounced increase of average weight gain compared to the other two groups, and Group 1 seem to have the lowest increase.

### Checking for outliers

```{r, warning=FALSE, message=FALSE}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time 1-64")
```

Each group has an outlier. Both Group 1 and Group 3 has an extreme low point, while Group 2 has an extreme high point. We can check from the data to find the outliers in each group. It also seems that the distribution of the second and third groups are quite skewed. Next, we will remove the outliers.

### Removing the outliers

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- filter(RATSL8S, (Group==1 & mean>250) | (Group==2 & mean < 590) | (Group==3 & mean>500))

# Draw a boxplot of the filtered data
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time 1-64")
```

After removing the outliers, the distribution of each group changed. This change is more apparent in the second and third group. The range of the quantiles seem to decrease quite significantly, and the skew of the distribution seem to have lessened.

### Statistical testing

We will not do a two-sample t-test as there are more than 2 groups in this data.

```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

The baseline is significant, which means that the original weight of the rats do affect the weight increase. On the other hand, Group is significant at 0.1 level, which means that we found a weak evidence that the diet difference affects the weight increase in rats.

## BPRS data

### Exploring the data

```{r, message=FALSE, warning=FALSE}
# loading the original data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)

# loading the wrangled data set
BPRSL <- read.csv(file = "bprsl.csv", sep=",", header=TRUE)

# Look at the (column) names 
names(BPRSL)

# Look at the structure 
str(BPRSL)

# Print out summaries of the variables
summary(BPRSL)
```

This data was obtained from 40 male subjects who were randomly assigned into 2 treatment groups. The subjects were then rated on the brief psychriatric rating scale (BPRS) which access various symptoms such as hostility, suspiciousness, and hallucination. The measurement was done before the treatment (week 0) and then weekly for eight weeks.

The wrangled long form data contains 360 rows in total and 5 variables which are:

| Variable  | Description                                       |
|-----------|---------------------------------------------------|
| treatment | Treatment group of each subject (integer, 1 or 2) |
| subject   | ID of each subject in each group (integer, 1-20)  |
| weeks     | Measurement week (character, 9 different entries) |
| bprs      | BPRS score (integer)                              |
| week      | Measurement week (integer, 0-8)                   |

### Factor the categorical variables

```{r}
# turn into factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# the data is already converted to the long form in the wrangling exercise

# checking 
glimpse(BPRSL)
```

The categorical variables had been successfully converted into factors.

### Plotting

```{r}
#mutating the subject variable for easier plotting
BPRSL$subject <- as.numeric(BPRSL$subject)
BPRSL <- mutate(BPRSL, subject = ifelse(treatment == "2", subject+20, subject))
BPRSL$subject <- factor(BPRSL$subject)

#plot
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(col = treatment))+ #changed linetype to col because it is easier to differentiate the groups with color
  scale_y_continuous(name = "BPRS")+
  theme(legend.position = "top")

```

In general, it seems that BPRS values decreases over time in general. However, it is not clear yet if there are any differences between the groups. The data in the second treatment group (blue) seem to have a wider range/variation compared to the first group.

### Create a regression model

```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs~ week+treatment, data=BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

The variable week is significant with a negative estimate, supporting the observation that BPRS score tend to decrease in time. However, the treatment variable is not found to be significant, which means that we did not find any evidence that the two treatments affects the BPRS score differently. It seems that the two treatments have a similar degree of impact to the BPRS score. We should note that this model ignores the independence assumption.

### Random intercept model

```{r, warning=FALSE, message=FALSE}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

This model allows the linear regression fit for each subject to differ in *intercept* from other subjects. The estimated standard deviation for subject in this model is quite large, which means that there are considerable differences in the intercept of each subject. Meanwhile, the estimates for the predictor variables are similar to the previous model.

### Random intercept and random slope model

```{r}
# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)
```

This model allows the linear regression fits for each subject to differ in both intercept and slope. This way, the differences in each subject's growth profiles and the effect of time can be both accounted for. In this model, the estimate for week is still similar, but the estimate for treatment is quite different. It seems that the difference of the treatment group is more apparent in this model compared to the previous ones.

### Anova test

```{r}
# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

The likelihood ratio test between the two models gives a chi square statistics of 63.663, and the associated p-value is very small. This means that the random intercept and slope model provides a better fit to the data.

### Interaction

```{r}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week*treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

In this model, we added an interaction between variables week and subject to the random slope random intercept model. Then, we compared this model to the previous one. The test between the two models produced a chi square value of 1.78 and the associated p-value is 0.18. As the p-value is quite large, it seems that the previous model (without interaction) is a better fit compared to this one (with interaction).

### Plotting the fitted values

```{r}
# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted 
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot with the fitted values
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line(aes(col = treatment))+
  scale_y_continuous(name = "BPRS")+
  theme(legend.position = "top")
```

This plot shows the fitted value for each subject according to the best fit model (random slope random intercept without interaction). We can see that in general, the BPRS score is decreasing for most of the subjects in both treatment groups, which may indicate that both treatment works in reducing BPRS score. However, we did not find enough evidence that the effect of two treatments differs.
